Using spatially biased variables in ecosystem condition accounting with a GIS based workflow

### 2024-02-22 manuscript/HIAs.qmd
test bolstad function

```{r terraToSTars}
#| eval: true
mire_stars_nf <- readRDS("manual_cache/mire_stars_nf.RDS")
mire_stars_gr <- readRDS("manual_cache/mire_stars_gr.RDS")
mire_stars_na <- readRDS("manual_cache/mire_stars_na.RDS")
```

@@

```{r muni-tbl}
#| eval: false

@@

# Take muni3 and add all sorts of other data to it
# using left_join.
# Save local cached file
muni_tbl <- muni3 |>
  # calculate total area
  mutate(
    area_km =
      round(
        units::drop_units(
          geom |> st_area() * 1e-6
        )
      )
  ) |>
  # make tibble for the coming join
  as_tibble() |>
  # paste inn terrestrial area
  left_join(
    terrestrial |>
      as_tibble() |>
      select(kommunenummer, t_area_km),
    keep = F
  ) |>
  # area of survey
  left_join(dk2 |> select(Municipality, dk_area_km)) |>
  mutate(dk_percent = round((dk_area_km / t_area_km) * 100)) |>
  # number of polygons
  left_join(nature3_tbl |> select(Municipality, n)) |>
  # total mire area and %
  left_join(mireArea |> select(Municipality, mirePercent, mire_km2)) |>
  # % mire inside survey coverage map
  left_join(mire_in_dk |> select(Municipality, mireInSurvey_km2)) |>
  mutate(mireInSurvery_percent = round(mireInSurvey_km2 / mire_km2 * 100, 2)) |>
  left_join(infraMuni3_summary)

saveRDS(muni_tbl, "../data/municipality_stats.RDS")
```

```{r read-munitbl}
#| eval: true
muni_tbl <- readRDS("../data/municipality_stats.RDS")
```

@@

| id  | Variable code        | Variable name                                    | Measurement unit                                                                        | Description                                                                                                                                                       | Reference                          |
|------------|------------|------------|------------|----------------|------------|
| 1   | 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                                                                                | @runehalvorsen2019                 |
| 2   | 7SE                  | Human caused abration or abration-caused erosion | Unitless, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 4 m^2^ quadrats layed over the area that has some sign of abration                                                             | @runehalvorsen2019                 |
| 3   | PRSL                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.                                        | Same as 7SE, but recorded at a higher resolution                                                                                                                  | @miljødirektoratet                 |
| 4   | 7TK                  | Tracks from large vehicles                       | Unitless, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 100 m^2^ quadrats layed over the area that has some signs of vehicle tracks                                                    | @rune_halvorsen_dokumentasjon_2019 |
| 5   | PRTK                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.                                        | Same as 7TK, but recorded at a higher resolution                                                                                                                  | @miljødirektoratet                 |
| 6   | 7GR-GI               | Ditching intensity                               | Unitless, ordinal scale from 1 to 5                                                     | Describes the effect that drainage ditches is estimate to have on the species composition and environmental variables ones the system reached its new equilibrium | @miljødirektoratet                 |
| 7   | Infrastructure Index | Infrastructure Index                             | Unitelss linear scale from 0 to 13.2                                                    | Unitelss index ranging from from 0 to 13.2                                                                                                                        | @erikstad_index_2023               |
| 8   | HIA                  | Human Impact area                                | Ordinal, non-linear scale from 1 to 4                                                   | A categorical representation of the infrastructure index                                                                                                          | this paper                         |

>>>>

```{r terraToSTars}
#| eval: true
mire_stars_nf <- readRDS("manual_cache/mire_stars_nf.RDS")
mire_stars_gr <- readRDS("manual_cache/mire_stars_gr.RDS")
mire_stars_na <- readRDS("manual_cache/mire_stars_na.RDS")
mire_terra_nf <- rast(mire_stars_nf)
mire_terra_gr <- rast(mire_stars_gr)
mire_terra_na <- rast(mire_stars_na)
```

@@

```{r muni-tbl}
#| eval: true
#| cache: true

@@

# Take muni3 and add all sorts of other data to it
# using left_join.
# Save local cached file
muni_tbl <- muni3 |>
  # calculate total area
  mutate(
    area_km =
      round(
        units::drop_units(
          geom |> st_area() * 1e-6
        )
      )
  ) |>
  # make tibble for the coming join
  as_tibble() |>
  # paste inn terrestrial area
  left_join(
    terrestrial |>
      as_tibble() |>
      select(kommunenummer, t_area_km),
    keep = F
  ) |>
  # area of survey
  left_join(dk2 |> select(Municipality, dk_area_km)) |>
  mutate(dk_percent = round((dk_area_km / t_area_km) * 100)) |>
  # number of polygons
  left_join(nature3_tbl |> select(Municipality, n)) |>
  # total mire area and %
  left_join(mireArea |> select(Municipality, mirePercent, mire_km2)) |>
  # % mire inside survey coverage map
  left_join(mire_in_dk |> select(Municipality, mireInSurvey_km2)) |>
  mutate(mireInSurvery_percent = round(mireInSurvey_km2 / mire_km2 * 100, 2)) |>
  left_join(infraMuni3_summary)
```

```{r read-munitbl}
#| eval: true
#muni_tbl <- readRDS("../data/municipality_stats.RDS")
```

# Now I need to spatially (horizontally) aggregate the indicator values for each 
# HIA. Rather than taking the arithmetic mean, I will use a Bayesian updating 
# approach. The point estimate (central tendency) will probably be the same
# more or less, with the two approaches, but with the updating approach I can
# get a honest measure for the uncertainty even with very small sample sizes.
# For this I need a 'true' value for the variation in the indicator
# I will use the entire national data set to determine this number

national_sd_ADSV <- sd(naturetypes$i_ADSV, na.rm=T)
national_sd_alien <- sd(naturetypes$i_alien, na.rm=T)
national_sd_ditch <- sd(naturetypes$i_ditch, na.rm=T)

barplot(c(national_sd_ADSV, national_sd_alien, national_sd_ditch),
        names.arg = c("ADSV", "Alien", "Trenching"),
        ylab="SD")

# The figure shows that the Trenching indicator is more spatially variable
```

```{r wgt-mean-fn}
#| eval: true

# This function is modified from Bolstad::normdp
# It updates a flat prior based on a sample of values (indicator values)
# and weight (polygon area) and returns a distribution for the weighted mean
# that has a gaussian distribution. It assumes the sampled population is also
# Gaussian. The variance is estimated from the full sample of indicator values.

wgt_mean <- function(x, weights, 
                     sigma.x = NULL, 
                     mu = seq(0, 1, length.out=1000), 
                     mu.prior = rep(1/length(mu), times=length(mu)), 
                     stat = "mean",
                     ...) {

  mx <- weighted.mean(x, weights)
  if (round(sum(mu.prior), 7) != 1) {
    warning("The prior probabilities did not sum to 1, therefore the prior has been normalized")
    mu.prior <- mu.prior / sum(mu.prior)
  }
  n.mu <- length(mu)
  nx <- length(x)
  snx <- sigma.x^2 / nx
  likelihood <- exp(-0.5 * (mx - mu)^2 / snx)
  posterior <- likelihood * mu.prior / sum(likelihood * mu.prior)
  mx <- sum(mu * posterior)
  vx <- sum((mu - mx)^2 * posterior)

  # draw 1k samples from the posterior to calculate the quantiles and sd from
  sample <- sample(mu, size = 1000, prob = posterior, replace = T)
  lower <- quantile(sample, probs = 0.025)
  upper <- quantile(sample, probs = 0.975)
  # we can assume that the distrubutin for the mean is gaussian, so we take a 
  # symetrical sd from here and use it, along with the mean, to recreate a
  # normal distribution later which we can sample from
  sdx <- sd(sample)
  results <- list(
    name = "mu", 
    param.x = mu, 
    prior = mu.prior,
    likelihood = likelihood, 
    posterior = posterior, 
    weighted_mean = mx,
    var = vx
  )
  if(stat == "mean") return(mx)
  if(stat == "lower") return(lower)
  if(stat == "upper") return(upper)
  if(stat == "sd") return(sdx)
}

```{r test-function}
#| eval: false
test_sd <- wgt_mean(x = nature3$i_ADSV[!is.na(nature3$i_ADSV)], 
                 weights = drop_units(nature3$area[!is.na(nature3$i_ADSV)]), 
                 sigma.x = national_sd_ADSV,
                 stat = "sd")
test_mean <- wgt_mean(x = nature3$i_ADSV[!is.na(nature3$i_ADSV)], 
                 weights = drop_units(nature3$area[!is.na(nature3$i_ADSV)]), 
                 sigma.x = national_sd_ADSV,
                 stat = "mean")

hist(rnorm(1000, test_mean, test_sd))

```{r meanPerHIA}
#| eval: true
#| cache: true

mu <- seq(0, 1, length.out=1000)
# Intersect nature3 with the HIA
temp <- nature3 |>
  st_intersection(infraMuni3) |>
  pivot_longer(cols = c(
    i_ADSV,
    i_alien,
    i_ditch),
    names_to = "indicator",
    values_to = "indicatorValue") |>
  filter(!is.na(indicatorValue)) |>
  group_by(indicator, Municipality, infrastructureIndex) |>
  summarise(mean )
```

@@

| id  | Variable code        | Variable name                                    | Measurement unit                                                                        | Description                                                                                                                                                       | Reference                          |
|----------|----------|----------|----------|-----------------------|----------|
| 1   | 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                                                                                | @runehalvorsen2019                 |
| 2   | 7SE                  | Human caused abration or abration-caused erosion | Unitless, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 4 m^2^ quadrats layed over the area that has some sign of abration                                                             | @runehalvorsen2019                 |
| 3   | PRSL                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.                                        | Same as 7SE, but recorded at a higher resolution                                                                                                                  | @miljødirektoratet                 |
| 4   | 7TK                  | Tracks from large vehicles                       | Unitless, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 100 m^2^ quadrats layed over the area that has some signs of vehicle tracks                                                    | @rune_halvorsen_dokumentasjon_2019 |
| 5   | PRTK                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.                                        | Same as 7TK, but recorded at a higher resolution                                                                                                                  | @miljødirektoratet                 |
| 6   | 7GR-GI               | Ditching intensity                               | Unitless, ordinal scale from 1 to 5                                                     | Describes the effect that drainage ditches is estimate to have on the species composition and environmental variables ones the system reached its new equilibrium | @miljødirektoratet                 |
| 7   | Infrastructure Index | Infrastructure Index                             | Unitelss linear scale from 0 to 13.2                                                    | Unitelss index ranging from from 0 to 13.2                                                                                                                        | @erikstad_index_2023               |
| 8   | HIA                  | Human Impact area                                | Ordinal, non-linear scale from 1 to 4                                                   | A categorical representation of the infrastructure index                                                                                                          | this paper                         |
